---
title: "Effect Size and Confidence Intervals in R - Helpful Tips"
author: "David Erceg-Hurn, Geoff Cumming, Robert Calin-Jageman"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Introductory stuff about book here

[R-Studio Resources for Learning R](https://www.rstudio.com/resources/training/online-learning/#R)
(with links to interactive tutorials for beginners)

[R Bloggers - How to Learn R](http://www.r-bloggers.com/how-to-learn-r-2)

[Quick-R](http://www.statmethods.net)

[Cookbook for R](http://www.cookbook-r.com/)

[R Quick Reference Card](http://cran.r-project.org/doc/contrib/Short-refcard.pdf)


## Picturing and Describing Data
Chapter 3 of ITNS.

Frequency Histogram
Stacked Dot Plot

For one, and two samples (side by side)

Descriptive Statistics
Mean, median, SD, quartiles, IQR, range, percentiles, z-scores

*Basic Descriptive Statistisc Functions*

Statistics          | Function
---------------------------------------
Mean                | mean()
Median              | median()
Minimum             | min()
Maximum             | max()
Standard Deviation  | sd()
Interquartile Range | IQR()
Range               | range()
Percentile          | quantile()

summary()





## The Normal Distribution and Sampling

Find tail areas for X or Z values
Find X given z

## Confidence Intervals and Effect Sizes

Cat's eye picture
Find tail areas for any T value, for any DF
Find critical value of t given df and a tail probability

Calculate MOE and 95% CI on a mean for single sample, usuing Z or t

## p Values, Null Hypothesis Significance Testing, and Confidence Intervals

Calculate p values gor single samples, using z or t

## Independent Groups Design
CI on difference between means
Cohen's d

## The Paired Design
CI on mean difference
d using SD-average (of pre and post) as standardizer


## Meta-Analysis

## Open Science and Planning Research

Any precision for planning functions
Any statistical power functions

## Correlation
Basic scatterplot
Adding a regression line

CIs on correlations
CI on difference between two independent correlations
Figure to display difference in two ind corr

## Regression
Calculationa nd figure for regression line in scatterplot

CI on the regression slope

Prediction intervals for individual values of Y at particular X values


## Frequencies, Proportions, and Risk
CI for a proportion
CI for difference between two independent proprtions
Figure for CI on a single proportion
Figure for difference with CI

ESCI uses Newcombe 1998 methods

Ratio between two variables - frequency tables
CI on the difference
Chi Square / Phi Coefficients

## Extended Designs - One Independent Variable
One way independent groups design
CI for planned contrasts, on two means
Figures if possible

One way repeated esigns, if possible (beyond scope of book and ESCI)

ANOVA p values

## Extended Designs - Two Indepenent Variables
ESCi only does 2 x 2 design
Means adn CI on mean effects
CI on single DF interaction (as difference in differences)

Simple man effect

Nice to include two way repeated measures if possible

## Open Science and Future Directions

Trimmed means for two independent means




Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
