---
title: "Getting Started with Effect Size and Confidence Intervals in R"
author: "David Erceg-Hurn, Geoff Cumming, Robert Calin-Jageman"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Introductory stuff about book here

[R-Studio Resources for Learning R](https://www.rstudio.com/resources/training/online-learning/#R)
(with links to interactive tutorials for beginners)

[R for Data Science](http://r4ds.had.co.nz/) - Online book by Garrett Grolemund and Hadley Wickham

[Data Camp have great interactive tutorials]
(https://www.datacamp.com/)

[Kelly Black's R Tutorial](http://www.cyclismo.org/tutorial/R/index.html)

[R Tutorial Series](https://rtutorialseries.blogspot.com.au/)

[R Bloggers - How to Learn R](http://www.r-bloggers.com/how-to-learn-r-2)

[Quick-R](http://www.statmethods.net)

[Cookbook for R](http://www.cookbook-r.com/)

[R Quick Reference Card](http://cran.r-project.org/doc/contrib/Short-refcard.pdf)

[inside-R](http://www.inside-r.org/)

Maybe effect size and CI stuff should be only thing in vignette, basic stuff like descriptive stats and plots taken out and put into separate file about installing and learning R on the itns website.


http://moc.environmentalinformatics-marburg.de/gitbooks/publicationQualityGraphics/_book/index.html  says:
google is your friend - for R related questions just type something like this 'r lattice how to change plot background color' The crux here is that you provide both the programming language - R - and the name of the package your question is related to - lattice. This way you will very likely find useful answers (the web is full of knowledgeable geeks)
for quick reference on the most useful basic functionality of R use http://www.statmethods.net
using google in the way outlined above, you will most likely end up at Stackoverflow at some stage. This is a very helpful platform for all sorts of programming issues with an ever increasing contribution of the R community. To search directly at Stackoverflow for R related stuff, type 'r' in front of you search.
Rseek is a search site dedicated exclusively to R-related stuff.
R-Bloggers is a nice site that provides access to all sorts of blog sites dedicated to R from all around the world (in fact, a lot of the material of this workshop is derived from posts found there).


## itns package
Describe how to install it here, say it is a collection of data sets for the book.  In rest of the vignette simply pointing users to useful functions and packages that they can use to do the analyses described in the book.

## Picturing and Describing Data
Chapter 3 of ITNS.

Frequency Histogram
Stacked Dot Plot

For one, and two samples (side by side)

Descriptive Statistics
Mean, median, SD, quartiles, IQR, range, percentiles, z-scores


### Basic Descriptive Statistics

All built into R.

Statistics          | Function
------------------- | -------------
Mean                | [mean()](http://www.inside-r.org/r-doc/base/mean)
Median              | [median()](http://www.inside-r.org/r-doc/stats/median)
Minimum             | [min()](http://www.inside-r.org/r-doc/base/Extremes)
Maximum             | [max()](http://www.inside-r.org/r-doc/base/Extremes)
Variance            | [var()](http://www.inside-r.org/r-doc/stats/cor)
Standard Deviation  | [sd()](http://www.inside-r.org/r-doc/stats/sd)
Interquartile Range | [IQR()](http://www.inside-r.org/r-doc/stats/IQR)
Range               | [range()](http://www.inside-r.org/r-doc/base/range)
Percentile          | [quantile()](http://www.inside-r.org/r-doc/stats/quantile)

[summary()](https://stat.ethz.ch/R-manual/R-devel/library/base/html/summary.html)

**Useful tutorials**
*[John Quirk's Summary and Descriptive Statistics](https://rtutorialseries.blogspot.com.au/2009/11/r-tutorial-series-summary-and.html)


#### Multiple groups
There are many ways to do this.  We recommend using dplyr package:

dplyr [vignette](https://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html), see the section on *Grouped Operations* using the group_by() and summarise() functions

Give an example of computing mean and standard deviation.

Other options include using the inbuilt R function [aggregate()](http://www.inside-r.org/r-doc/stats/aggregate) or the [doBy package](https://cran.r-project.org/web/packages/doBy/vignettes/doBy.pdf).


### Z-Scores
[John Quick's tutorial](http://www.r-bloggers.com/r-tutorial-series-centering-variables-and-generating-z-scores-with-the-scale-function) shows how to use R's inbuilt scale() function to compute Z-scores.  See also [Seam Dolinar's
tutorial](http://stats.seandolinar.com/calculating-z-scores-with-r/) on Calculating Z scores and finding tail probabilities.

## Plots
Three graphics systems - base graphics, lattice, and ggplot2.

For new users, would recommend learning ggplot2 first as it is the most flexible.

Deducer has an excellent GUI for ggplot2 that makes it easier to create plots.

[This tutoral](http://personal.colby.edu/personal/m/mgimond/RIntro/05_Plotting_systems.html) provides an overview of the three systems with examples.

Tim Appelhans has a [comprehensive tutorial](http://moc.environmentalinformatics-marburg.de/gitbooks/publicationQualityGraphics/_book/chapters/02_data_visualisation/scatter_lattice.html) about creating publication quality graphics using lattice and ggplot2 systems.

he also notes that
"google is your friend - for R related questions just type something like this 'r lattice how to change plot background color' The crux here is that you provide both the programming language - R - and the name of the package your question is related to - lattice. This way you will very likely find useful answers (the web is full of knowledgeable geeks)""


Histogram

Dotplot



Best place to start is probably the [Getting Started with Lattice Graphics](http://lattice.r-forge.r-project.org/Vignettes/src/lattice-intro/lattice-intro.pdf) tutorial by Deepayan Sarkar.

His [Lattice book](https://www.springer.com/gp/book/9780387759685) provides more comprehensive information.


[ggplot2 book website](http://ggplot2.org/book/)
[ggplot2 online documentation](http://docs.ggplot2.org/current/index.html)

[Winston Chang's R Graphic's Cookbook](http://www.cookbook-r.com/Graphs)

[Harvard Introduction to ggplot2](http://tutorials.iq.harvard.edu/R/Rgraphics/Rgraphics.html)

[Zevross ggplot2 tutoial](http://zevross.com/blog/2014/08/04/beautiful-plotting-in-r-a-ggplot2-cheatsheet-3/)

[ggplot2 tutorial](https://github.com/jennybc/ggplot2-tutorial)

DataCamp online courses (cost money)
https://www.datacamp.com/courses/data-visualization-with-ggplot2-1

[RStudio ggplot2 cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)

## Confidence Intervals and Effect Sizes

Cat's eye picture
Find tail areas for any T value, for any DF
Find critical value of t given df and a tail probability

Calculate MOE and 95% CI on a mean for single sample, usuing Z or t

## The Normal Distribution and Sampling
Find tail areas for X or Z values
Find X given z

## p Values, Null Hypothesis Significance Testing, and Confidence Intervals

Calculate p values gor single samples, using z or t

[Kelly Black's tutorial](http://www.cyclismo.org/tutorial/R/pValues.html) shows how to compute p values using z- or t-distributions.

[Kelly Black's tutorial](http://www.cyclismo.org/tutorial/R/confidence.html)  shows how to calculate confidence intervals for means using normal or t-distributions.

The easiest way to compute confidence intervals and p values is using the t.test() function.

ggplot showing CI on a mean?

## t.test() function
The t.test() function is built into R.  It produces confidence intervals and p-values for single samples, two independent groups, and paired samples.

Give examples using itns data.

t.test(x, y, conf.level=ci, var.equal = TRUE)
t.test(x, y, conf.level=ci, var.equal = FALSE)
t.test(x, y, paired = TRUE)

## MBESS package
library(MBESS)
Ken Kelly's [MBESS (Methods for the Behavioural and Social Sciences)](https://www3.nd.edu/~kkelley/site/MBESS.html) package.

Includes functions for computing confidence intervals around the standardised mean difference, unstandardised and standardised regression coefficients, reliability coefficents, ANOVA contrasts, R-squared, and more.  Also includes functions for power analysis and sample size planning.

Demonstrate some key functions related to book.

CI on difference between means
Cohen's d

Effect Sizes and Confidence Intervals
Sample Size and Parameter Estimation Sample Size Planning

Any precision for planning functions
Any statistical power functions

## WRS and WRS2 packages



## Standardized Paired Difference
Is there any function for this?
Or show how to compute by hand and use noncentral T from MBESs
CI on mean difference
d using SD-average (of pre and post) as standardizer


## Meta-Analysis
There are numerous R packages that can be used to conduct meta-analyses for a wide variety of effect sizes such as means, mean differences, standardized mean differences, proportions, odds rations, etc. See the [CRAN Meta-Analysis Task View](https://cran.r-project.org/web/views/MetaAnalysis.html) for a comprehensive list of them.

A popular and well documented package for conducting meta-analyses in R is [metafor](http://www.metafor-project.org).  See the comprehensive [metafor website](http://www.metafor-project.org) for more information.

[metagear](http://lajeunesse.myweb.usf.edu/publications.html) 
is a relatively new package which has meta-analytic capabilities as well as functions that help users conduct systematic reviews  and generate [PRISMA (Preferred Reporting Items for Systematic Reviews and Meta-Analyses)](http://www.prisma-statement.org/) flow charts.  [This vignette](http://lajeunesse.myweb.usf.edu/metagear/metagear_basic_vignette.html) provides an overview of the metagrear package.

Other useful sources of information about conducting meta-analyses in R include:  
* [A.C Del Re's Practical Tutorial](http://www.tqmp.org/RegularArticles/vol11-1/p037/p037.pdf) on Conducting Meta-Analysis in R using the metafor and MAd packages
* Stephanie Kovalchik's [Tutorial on Meta-Analysis in R](http://www.edii.uclm.es/~useR-2013/Tutorials/kovalchik/kovalchik_meta_tutorial.pdf) from the 2013 useR! Conference
* Schwarzer, Carpenter, and Rucker's [Meta-Analysis with R](https://link.springer.com/book/10.1007/978-3-319-21416-0) book
* [R-Studio's tutorial](https://rstudio-pubs-static.s3.amazonaws.com/10913_5858762ec84b458d89b0f4a4e6dd5e81.html) on running meta-analyses in R using the metafor package
* Simon Knight's Guide to Meta-Analysis in R - [part 1](http://www.transplantevidence.com/blog/2014/07/meta-analysis-in-r-part-1-installing-the-software/) and [part 2](http://www.transplantevidence.com/blog/2014/08/meta-analysis-in-r-part-2-binary-data).
* Stephanie Hick's [Easy Introduction to Meta-Analysis in R](https://statisticalrecipes.blogspot.com.au/2014/01/easy-introduction-to-meta-analyses-in-r.html) using the meta package 


## multicon plots
egraph() - A function for plotting means as dots and error bars without caps around them
catseye() - A function for creating cat's eye plots of group means
diffPlot() - A function for creating difference plots for two group (both paired and independent) comparisons


## Open Science and Planning Research

MBESs does this

## Correlation
Basic scatterplot
Adding a regression line

CIs on correlations
CI on difference between two independent correlations
Figure to display difference in two ind corr

## Regression
Calculation and figure for regression line in scatterplot

CI on the regression slope

lm() is the inbuilt R function for ordinary least-squares regression.
confint() gives confidence intervals on regression parameters

Also MBESS can do this (apaprently) as well as for standardized coefficients.

Link to graphics.

Show ggplot2 scatterplot with OLS line, maybe also show smoother, mention it can also fit robust and other regression lines.

Also countless other packages for fitting a variety of regression models in R.

Prediction intervals for individual values of Y at particular X values


## Frequencies, Proportions, and Risk
CI for a proportion
CI for difference between two independent proprtions
Figure for CI on a single proportion
Figure for difference with CI

ESCI uses Newcombe 1998 methods

Ratio between two variables - frequency tables
CI on the difference
Chi Square / Phi Coefficients

vcd and vcdExtra packages for visualising categorical data

PropCIs
Does CIs for single, independent, and paired proportions.
Includes risk score CI.

library(pairwiseCI)	# for ARR NHS and RR using Score Method.  Uses Prop.diff and Prop.ratio	

And chi-square function I used for recent WCBCT analyses.

R manual to accompany Agresti's Categorical Data Analysis 2nd Ed by Laura Thompson
http://www.stat.ufl.edu/~aa/cda/Thompson_manual.pdf

## Extended Designs - One Independent Variable
One way independent groups design
CI for planned contrasts, on two means
Figures if possible

One way repeated esigns, if possible (beyond scope of book and ESCI)

ANOVA p values

## Extended Designs - Two Indepenent Variables
ESCi only does 2 x 2 design
Means adn CI on mean effects
CI on single DF interaction (as difference in differences)

Simple man effect

Nice to include two way repeated measures if possible

## Open Science and Future Directions

Trimmed means for two independent means




Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
